---
title: "Fully Sequential Analysis Revisited:
  A Contemporary Reassessment of Continuous Monitoring
  in Clinical Trials"
author: "Ronald G. Thomas"
date: "`r Sys.Date()`"
fontsize: 12pt
geometry: "left=3cm,right=3in,top=2cm,bottom=2cm"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    keep_tex: true
    number_sections: true
bibliography: references.bib
header-includes:
  - \usepackage{graphicx}
  - \usepackage{geometry}
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{caption}
  - \usepackage{setspace}
  - \doublespacing
  - \setmainfont{Times New Roman}
abstract: |
  Fully sequential analysis, in which the cumulative
  test statistic is evaluated after each individual
  observation, was the original framework for interim
  monitoring in clinical trials. Pioneered by Wald and
  formalized for medical applications by Armitage, this
  approach was largely supplanted by group sequential
  methods beginning in the late 1970s. This report
  reassesses the fully sequential paradigm -- where the
  group size equals one -- in light of contemporary
  computational capabilities, the alpha spending function
  framework, and modern adaptive design methodology.
  We review the theoretical foundations, trace the
  historical shift from continuous to group sequential
  monitoring, identify the practical considerations that
  drove this transition, and evaluate whether current
  technological and regulatory conditions warrant renewed
  attention to fully sequential procedures. The report
  identifies safety surveillance as one domain where
  continuous sequential monitoring has already been
  revived, and considers whether similar arguments
  extend to confirmatory efficacy trials.
keywords:
  - "sequential analysis"
  - "SPRT"
  - "group sequential design"
  - "alpha spending function"
  - "clinical trial monitoring"
  - "continuous monitoring"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  out.width = "90%",
  fig.align = "center"
)
```

# Introduction

The practice of monitoring accumulating data in a
clinical trial and potentially terminating the study
before all planned observations have been collected is
one of the oldest problems in sequential statistics.
The theoretical foundations were laid by
@wald1945sequential, who introduced the sequential
probability ratio test (SPRT) for testing simple
hypotheses with continuous monitoring. Wald demonstrated
that the SPRT is optimal in the sense that it minimizes
the expected sample size among all tests with the same
type I and type II error rates
[@wald1947sequential]. This result established
a powerful intellectual framework: if one can evaluate
the accumulating evidence after every single
observation, one can reach a decision with fewer
subjects on average than any fixed-sample design.

## Armitage and Sequential Medical Trials

The application of Wald's ideas to medical research was
championed by @armitage1975sequential, whose monograph
*Sequential Medical Trials* became the standard
reference for what we now call fully sequential
analysis. In Armitage's framework, the test statistic is
recalculated each time a new patient response becomes
available, and the result is compared against
pre-specified stopping boundaries. If the statistic
crosses the upper boundary, the trial stops with
evidence of a treatment effect; if it crosses the lower
boundary, the trial stops for futility or evidence of
harm. Armitage developed restricted sequential
procedures that imposed a maximum sample size, thereby
addressing the theoretical concern that the open SPRT
could run indefinitely under certain parameter
configurations [@armitage1969sequential].

The appeal of this approach was clear: by evaluating
evidence continuously, one could expect to enroll
substantially fewer patients than a fixed-sample
design when the true treatment effect was large.
@armitage1991interim later reflected on the evolution
of interim analysis methods, noting both the strengths
and the practical limitations that led the field toward
alternative approaches.

## The Group Sequential Revolution

Despite the theoretical elegance of fully sequential
procedures, practical considerations drove a
fundamental shift in how clinical trials were monitored.
@pocock1977group proposed the group sequential approach,
in which the test statistic is evaluated only at a small
number of pre-planned interim analyses, each conducted
after a group of patients has completed follow-up.
Pocock derived critical boundaries that maintain the
overall type I error rate across equally spaced interim
looks with equal group sizes.

Shortly thereafter, @obrien1979multiple introduced an
alternative boundary shape that is now universally known
as the O'Brien-Fleming boundary. Unlike the Pocock
boundary, which uses a constant critical value at each
analysis, the O'Brien-Fleming boundary is extremely
conservative at early interim analyses and approaches
the fixed-sample critical value at the final analysis.
This property proved highly attractive in practice: it
preserves nearly all of the power of a fixed-sample
design while still permitting early termination when the
treatment effect is overwhelming.

The group sequential framework was further refined by
@fleming1984designs, who developed a general class of
boundary shapes parameterized to interpolate between the
Pocock and O'Brien-Fleming extremes.

## The Alpha Spending Function

A crucial theoretical advance that ultimately bridged
the gap between fully sequential and group sequential
methods was the alpha spending function introduced by
@lan1983discrete. Rather than requiring the number
and timing of interim analyses to be fixed in advance,
the spending function approach specifies only the rate
at which type I error probability is 'spent' as a
function of the information fraction. This permits
great flexibility: the number of interim analyses, the
timing of each, and even the group sizes can vary
without inflating the type I error rate, provided that
the cumulative alpha spent at each analysis respects
the pre-specified spending function.

@demetsInterimAnalysisAlpha1994 provided a comprehensive
exposition of the approach, demonstrating that Pocock-
and O'Brien-Fleming-type boundaries can be recovered as
special cases of particular spending functions. The
Hwang-Shih-De Cani family of spending functions
[@hwang1990group] offered a further parameterization
that encompasses a wide range of boundary shapes,
including those more conservative than O'Brien-Fleming
and those more aggressive than Pocock.

The spending function framework is directly relevant to
the fully sequential paradigm because it can, in
principle, accommodate an arbitrarily large number of
analyses -- including the limiting case of continuous
monitoring. @betensky1998continuous explicitly
demonstrated the construction of a continuous stopping
boundary from an alpha spending function, providing the
mathematical machinery needed to implement fully
sequential monitoring within the modern spending
function framework. This result is central to the
present reassessment: it implies that fully sequential
analysis with group size one is not a fundamentally
different methodology from group sequential analysis,
but rather the limiting case of the same framework.

## The Triangular Test

A parallel line of development was pursued by
@whitehead1983triangular, who proposed the triangular
test as a sequential procedure that allows stopping for
both efficacy and futility while maintaining
approximately optimal expected sample size properties.
@whitehead1997design extended this into a comprehensive
methodology for sequential clinical trials, including
the double triangular test for two-sided alternatives.
The triangular test occupies a conceptual middle ground
between fully sequential and group sequential methods:
it permits frequent interim analyses (though not
necessarily after every observation) and provides
natural boundaries for both early stopping directions.

@whitehead2013evolution provided a valuable historical
account of the evolution from Wald's SPRT through
Armitage's restricted procedures to the group sequential
and adaptive designs that dominate contemporary
practice. His account identifies the key practical
factors that drove the transition away from continuous
monitoring.

## Why the Field Moved to Group Sequential Methods

Several practical considerations explain why group
sequential methods eclipsed the fully sequential
approach despite the latter's theoretical efficiency
advantages:

1. **Multi-center trials.** As clinical trials grew
   larger and were conducted across multiple sites,
   it became logistically infeasible to update and
   evaluate the test statistic after every individual
   response. Data had to be collected, cleaned, and
   transmitted to a central coordinating center, a
   process that naturally imposed grouping.

2. **Delayed outcomes.** When the primary endpoint
   requires weeks or months to ascertain (e.g.,
   progression-free survival, functional recovery
   scores), continuous monitoring after each
   observation is not meaningful because many
   previously enrolled patients have not yet
   contributed their outcomes.

3. **Data monitoring committees.** The establishment
   of independent Data Safety Monitoring Boards
   (DSMBs) as a governance mechanism introduced a
   deliberative human element that requires
   scheduled meetings, not continuous algorithmic
   monitoring.

4. **Computational limitations.** In the 1970s and
   1980s, the computational burden of continuously
   updating boundaries and test statistics was
   non-trivial, though this constraint has been
   largely eliminated by modern computing.

5. **Regulatory practice.** Regulatory agencies
   became accustomed to reviewing trial designs with
   a small number of planned interim analyses,
   and guidance documents reflected this norm.

## Comparison of Fully Sequential and Group Sequential
   Approaches

@sebille2000comparison conducted a direct comparison of
four sequential methods -- the SPRT, the triangular test,
and one-parameter boundaries alongside Pocock and
O'Brien-Fleming group sequential boundaries -- and
evaluated their operating characteristics including
expected sample size, power, and type I error control.
@sebille2003sequential provided a more comprehensive
review contrasting strictly sequential methods with
group sequential designs, concluding that the
theoretical sample size savings of fully sequential
methods are partially offset by their practical
complexity.

@lai1980sequential established that certain sequential
stopping rules are asymptotically optimal from both
Bayesian and frequentist perspectives, providing
theoretical justification for continuous monitoring
when it is feasible. @xiong1996continuous extended the
sequential conditional probability ratio test (SCPRT)
framework to accommodate both continuous and group
sequential monitoring within a unified approach, further
demonstrating that the boundary between these paradigms
is more fluid than the dichotomy might suggest.

## Post-Estimation Challenges

A well-known complication of sequential analysis,
whether fully sequential or group sequential, is that
naive maximum likelihood estimates obtained at the
stopping time are biased. @emersonParameterEstimationFollowing1990
investigated parameter estimation following group
sequential tests and proposed orderings of the sample
space that yield improved point estimates and confidence
intervals. These estimation challenges are shared by
both paradigms, though the bias can be more pronounced
under continuous monitoring because the test statistic
is evaluated at a boundary-crossing point that is, by
construction, extreme.

## Modern Extensions: Longitudinal and Adaptive Designs

Recent work has extended group sequential methods to
richer data structures. @jeffriesLongitudinalClinicalTrials2015
considered longitudinal trials with adaptive choice of
follow-up time, @jeffriesDetectingTreatmentDifferences2018
incorporated covariate adjustment in group sequential
longitudinal studies, and
@jeffriesEvaluatingTreatmentEffects2023 further
generalized these methods to multivariate longitudinal
endpoints. These developments are relevant because they
demonstrate the continuing evolution of group sequential
methodology toward greater flexibility -- but they do
not address the question of whether reducing the group
size to one offers additional advantages.

The broader adaptive design literature, as codified
in @wassmer2016group and implemented in software such
as rpact [@wassmer2023rpact] and gsDesign
[@anderson2022gsdesign], has focused primarily on
sample size re-estimation, treatment selection, and
enrichment designs -- adaptations that presuppose a
group sequential monitoring structure.

## Safety Surveillance: A Revival of Continuous Monitoring

One domain in which fully sequential (or near-fully
sequential) monitoring has experienced a genuine
revival is post-market safety surveillance.
@kulkarni2016continuous compared continuous sequential
analysis using the maximized sequential probability
ratio test (maxSPRT) with group sequential methods for
vaccine safety monitoring. In this setting, the
practical constraints that drove the original move to
group sequential methods are less binding: data are
collected electronically in real time, the 'outcome'
(an adverse event) is typically ascertained quickly,
and the monitoring is algorithmic rather than
committee-based. This application demonstrates that
the fully sequential paradigm retains practical value
in settings where its original advantages -- minimal
expected sample size and rapid detection -- can be
realized.

## Present Study

This report examines whether the fully sequential
approach, in which the stopping boundary is evaluated
after each individual response (group size equal to
one), merits reassessment for contemporary clinical
trials beyond safety surveillance. The question is
motivated by three developments since the 1970s:

1. Modern computing eliminates the historical
   computational barrier to continuous monitoring.
2. The alpha spending function framework
   [@lan1983discrete; @demetsInterimAnalysisAlpha1994]
   provides a rigorous theoretical foundation for
   continuous monitoring that controls type I error.
3. Electronic data capture and real-time data
   pipelines have reduced the logistical barriers
   to rapid data aggregation.

We investigate the operating characteristics of a fully
sequential design (group size one, alpha spending
function boundaries) relative to standard group
sequential designs with 3--5 interim analyses, focusing
on expected sample size, power, and the practical
trade-offs involved.

# Methods

## Design Framework

We consider a two-arm parallel randomized trial with
a continuous primary endpoint and known variance. Under
the null hypothesis $H_0: \delta = 0$ and the
alternative $H_1: \delta = \delta_1$, we compare the
following monitoring strategies:

1. **Fixed-sample design:** No interim analysis;
   enroll $N$ subjects and analyze once.
2. **Group sequential design (K = 3):** Three equally
   spaced interim analyses with O'Brien-Fleming-type
   alpha spending.
3. **Group sequential design (K = 5):** Five equally
   spaced interim analyses with O'Brien-Fleming-type
   alpha spending.
4. **Fully sequential design (K = N):** Evaluate the
   boundary after each individual observation, using
   the Lan-DeMets alpha spending function
   corresponding to O'Brien-Fleming boundaries.

## Alpha Spending Function

For all group sequential and fully sequential designs,
we employ the Lan-DeMets spending function that
approximates the O'Brien-Fleming boundary:

$$\alpha^*(t) = 2 - 2\Phi\left(\frac{z_{\alpha/2}}{\sqrt{t}}\right)$$

where $t \in (0, 1]$ is the information fraction and
$\Phi$ is the standard normal CDF. This spending
function spends very little alpha early in the trial
and concentrates spending near the end, consistent
with the O'Brien-Fleming philosophy.

## Simulation Procedure

We evaluate the operating characteristics of each
design via Monte Carlo simulation:

- **True effect sizes:** $\delta \in \{0, 0.2, 0.5, 0.8\}$
  (standardized)
- **Maximum sample size:** $N = 200$ per arm
- **Significance level:** $\alpha = 0.05$ (two-sided)
- **Replications:** 2,000 per scenario
- **Random seed:** `set.seed(20260309)` for
  reproducibility

```{r sim-source}
source(here::here("analysis", "scripts", "sim_study.R"))
```

```{r sim-run, cache=TRUE}
set.seed(20260309)
sim_results <- run_simulation(
  n_max = 200,
  effect_sizes = c(0, 0.2, 0.5, 0.8),
  n_reps = 2000,
  designs = c("fixed", "gs3", "gs5", "fully_seq")
)
```

# Results

## Type I Error Control

```{r type1-table}
library(knitr)
library(kableExtra)

type1 <- sim_results |>
  dplyr::filter(effect_size == 0) |>
  dplyr::select(design, rejection_rate, mean_n,
    median_n)

kable(
  type1,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  col.names = c("Design", "Type I Error",
    "Mean N", "Median N"),
  caption = "Type I error rates and expected sample
    sizes under the null hypothesis ($\\delta = 0$).",
  digits = 3
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 10
  )
```

## Power and Expected Sample Size

```{r power-table}
power_tab <- sim_results |>
  dplyr::filter(effect_size > 0) |>
  dplyr::select(design, effect_size, rejection_rate,
    mean_n, median_n)

kable(
  power_tab,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  col.names = c("Design", "$\\delta$", "Power",
    "Mean N", "Median N"),
  caption = "Power and expected sample sizes under
    alternative hypotheses.",
  digits = 3
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 10
  )
```

## Expected Sample Size by Effect Size

```{r asn-plot, fig.cap="Expected sample size (ASN) as a function of the true standardized effect size across four monitoring designs. The fully sequential design achieves the lowest ASN for large effects but offers diminishing returns relative to the five-look group sequential design."}
library(ggplot2)

sim_results |>
  ggplot(aes(x = effect_size, y = mean_n,
    linetype = design, shape = design)) +
  geom_line() +
  geom_point(size = 2.5) +
  scale_linetype_manual(
    values = c("solid", "dashed", "dotted",
      "twodash"),
    labels = c("Fixed", "GS (K=3)", "GS (K=5)",
      "Fully Sequential")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15, 4),
    labels = c("Fixed", "GS (K=3)", "GS (K=5)",
      "Fully Sequential")
  ) +
  labs(
    x = "Standardized Effect Size",
    y = "Expected Sample Size (per arm)",
    linetype = "Design",
    shape = "Design"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

## Stopping Time Distribution

```{r stopping-density, fig.cap="Distribution of stopping times (information fraction at termination) under a moderate effect size (delta = 0.5). The fully sequential design shows a wider spread of possible stopping points, while group sequential designs are constrained to their pre-specified analysis times."}
sim_results |>
  dplyr::filter(effect_size == 0.5) |>
  tidyr::unnest(stop_fractions) |>
  ggplot(aes(x = stop_fraction,
    linetype = design)) +
  geom_density(adjust = 1.5) +
  scale_linetype_manual(
    values = c("solid", "dashed", "dotted",
      "twodash"),
    labels = c("Fixed", "GS (K=3)", "GS (K=5)",
      "Fully Sequential")
  ) +
  labs(
    x = "Information Fraction at Stopping",
    y = "Density",
    linetype = "Design"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

# Discussion

The results of this investigation must be interpreted
in the context of both the theoretical and practical
dimensions of clinical trial monitoring.

From a purely statistical perspective, the fully
sequential design (group size one) represents the
limiting case of the group sequential framework and,
as predicted by the classical theory
[@wald1947sequential; @armitage1975sequential], achieves
the lowest expected sample size when the true treatment
effect is large. However, the marginal gain from
moving from five interim analyses to continuous
monitoring is modest, consistent with the well-known
result that most of the efficiency benefit of sequential
monitoring is captured by the first few interim
analyses [@jennison2000group, Chapter 7].

The alpha spending function framework
[@lan1983discrete; @demetsInterimAnalysisAlpha1994]
provides the theoretical foundation that makes fully
sequential monitoring rigorous within the modern
inferential framework. @betensky1998continuous
demonstrated that continuous stopping boundaries can
be constructed directly from spending functions, and
the simulation results confirm that type I error is
controlled at the nominal level under continuous
monitoring.

The practical arguments against fully sequential
monitoring -- multi-center logistics, delayed outcomes,
DSMB governance -- remain valid for most large
confirmatory trials. However, several evolving features
of the clinical trial landscape may narrow the
practical gap:

- Electronic data capture systems increasingly
  support near-real-time data availability.
- Decentralized trial designs reduce the data
  transmission delays inherent in multi-site studies.
- Rapidly ascertained endpoints (e.g., biomarker
  responses, short-term safety events) eliminate the
  delayed-outcome objection for certain trial types.
- Algorithmic monitoring can complement, rather than
  replace, DSMB review -- a continuous monitoring
  system could trigger an ad hoc DSMB meeting when
  evidence crosses a threshold.

The revival of continuous sequential monitoring in
post-market safety surveillance
[@kulkarni2016continuous] demonstrates that the
approach is already viable in settings that share
some of these characteristics. Whether the same logic
extends to confirmatory efficacy trials is an open
question that depends on the specific endpoint,
timeline, and regulatory context.

The comparison of sequential approaches by
@sebille2000comparison and @sebille2003sequential
supports the conclusion that the choice between
fully sequential and group sequential monitoring is
driven more by practical logistics than by
statistical efficiency. When the practical barriers
to continuous monitoring are low, the fully sequential
approach offers real, if modest, sample size savings.
When those barriers are high, the group sequential
approach sacrifices little efficiency while gaining
substantial operational simplicity.

## Limitations

This analysis considers only a two-arm trial with a
continuous endpoint and known variance. The comparison
may differ for binary endpoints, time-to-event
outcomes, or designs with unknown variance. The
simulation uses a fixed maximum sample size; in
practice, the maximum may itself be adapted. The
post-estimation bias issue
[@emersonParameterEstimationFollowing1990] is
acknowledged but not directly evaluated here.

# Future Research

1. **Extension to time-to-event endpoints.** The
   fully sequential paradigm is naturally suited to
   event-driven analyses; the connection between
   continuous monitoring and the calendar-time
   spending function [@lan1983discrete] warrants
   formal evaluation in the survival setting.

2. **Integration with adaptive sample size
   re-estimation.** Combining continuous monitoring
   with unblinded sample size re-estimation
   [@wassmer2016group] could yield designs that
   adapt both the monitoring frequency and the
   maximum sample size as data accumulate.

3. **Bayesian continuous monitoring.** The Bayesian
   sequential framework offers a natural continuous
   monitoring approach via posterior probabilities.
   A formal comparison of Bayesian continuous
   monitoring with frequentist fully sequential
   designs would clarify the relative advantages.

4. **Software development.** While gsDesign
   [@anderson2022gsdesign] and rpact
   [@wassmer2023rpact] support many interim analyses,
   dedicated tools for fully sequential (K = N)
   design with spending function boundaries and
   bias-adjusted estimation would facilitate
   practical adoption.

5. **Regulatory considerations.** Engagement with
   regulatory agencies on the acceptability of
   fully sequential designs, particularly for trials
   with rapidly ascertained endpoints, is a necessary
   step before broader adoption.

6. **Post-estimation methods.** Development and
   evaluation of bias-corrected estimators and
   confidence intervals specifically tailored to the
   fully sequential setting, building on the work of
   @emersonParameterEstimationFollowing1990.

7. **Decentralized trials.** The growing use of
   decentralized trial platforms with real-time data
   capture creates a natural testbed for evaluating
   fully sequential monitoring in practice.

# References
