---
title: "Fully Sequential Analysis Revisited:
  A Contemporary Reassessment of Continuous Monitoring
  in Clinical Trials"
author: "Ronald G. Thomas, Ph.D.\\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
abstract: |
  Fully sequential analysis, in which the cumulative
  test statistic is evaluated after each individual
  observation, was the original framework for interim
  monitoring in clinical trials. Pioneered by Wald and
  formalized for medical applications by Armitage, this
  approach was largely supplanted by group sequential
  methods beginning in the late 1970s. This report
  reassesses the fully sequential paradigm -- where the
  group size equals one -- in light of contemporary
  computational capabilities, the alpha spending function
  framework, and modern adaptive design methodology.
  We review the theoretical foundations, trace the
  historical shift from continuous to group sequential
  monitoring, identify the practical considerations that
  drove this transition, and evaluate whether current
  technological and regulatory conditions warrant renewed
  attention to fully sequential procedures. The report
  identifies safety surveillance as one domain where
  continuous sequential monitoring has already been
  revived, and considers whether similar arguments
  extend to confirmatory efficacy trials.
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
linestretch: 1.1
geometry: 'left=3cm,right=3cm,top=2.5cm,bottom=2.5cm'
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage[mathlines]{lineno}
  - \linenumbers
  - \usepackage{graphicx}
  - \usepackage{caption}
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  out.width = "90%",
  fig.align = "center"
)
```

# Introduction

The practice of monitoring accumulating data in a
clinical trial and potentially terminating the study
before all planned observations have been collected is
one of the oldest problems in sequential statistics.
The theoretical foundations were laid by
@wald1945sequential, who introduced the sequential
probability ratio test (SPRT) for testing simple
hypotheses with continuous monitoring. Wald demonstrated
that the SPRT is optimal in the sense that it minimizes
the expected sample size among all tests with the same
type I and type II error rates
[@wald1947sequential]. This result established
a powerful intellectual framework: if one can evaluate
the accumulating evidence after every single
observation, one can reach a decision with fewer
subjects on average than any fixed-sample design.

## Armitage and Sequential Medical Trials

The application of Wald's ideas to medical research was
championed by @armitage1975sequential, whose monograph
*Sequential Medical Trials* became the standard
reference for what we now call fully sequential
analysis. In Armitage's framework, the test statistic is
recalculated each time a new patient response becomes
available, and the result is compared against
pre-specified stopping boundaries. If the statistic
crosses the upper boundary, the trial stops with
evidence of a treatment effect; if it crosses the lower
boundary, the trial stops for futility or evidence of
harm. Armitage developed restricted sequential
procedures that imposed a maximum sample size, thereby
addressing the theoretical concern that the open SPRT
could run indefinitely under certain parameter
configurations [@armitage1969sequential].

The appeal of this approach was clear: by evaluating
evidence continuously, one could expect to enroll
substantially fewer patients than a fixed-sample
design when the true treatment effect was large.
@armitage1991interim later reflected on the evolution
of interim analysis methods, noting both the strengths
and the practical limitations that led the field toward
alternative approaches.

## The Group Sequential Revolution

Despite the theoretical elegance of fully sequential
procedures, practical considerations drove a
fundamental shift in how clinical trials were monitored.
@pocock1977group proposed the group sequential approach,
in which the test statistic is evaluated only at a small
number of pre-planned interim analyses, each conducted
after a group of patients has completed follow-up.
Pocock derived critical boundaries that maintain the
overall type I error rate across equally spaced interim
looks with equal group sizes.

Shortly thereafter, @obrien1979multiple introduced an
alternative boundary shape that is now universally known
as the O'Brien-Fleming boundary. Unlike the Pocock
boundary, which uses a constant critical value at each
analysis, the O'Brien-Fleming boundary is extremely
conservative at early interim analyses and approaches
the fixed-sample critical value at the final analysis.
This property proved highly attractive in practice: it
preserves nearly all of the power of a fixed-sample
design while still permitting early termination when the
treatment effect is overwhelming.

The group sequential framework was further refined by
@fleming1984designs, who developed a general class of
boundary shapes parameterized to interpolate between the
Pocock and O'Brien-Fleming extremes.

## The Alpha Spending Function

A crucial theoretical advance that ultimately bridged
the gap between fully sequential and group sequential
methods was the alpha spending function introduced by
@lan1983discrete. Rather than requiring the number
and timing of interim analyses to be fixed in advance,
the spending function approach specifies only the rate
at which type I error probability is 'spent' as a
function of the information fraction. This permits
great flexibility: the number of interim analyses, the
timing of each, and even the group sizes can vary
without inflating the type I error rate, provided that
the cumulative alpha spent at each analysis respects
the pre-specified spending function.

@demetsInterimAnalysisAlpha1994 provided a comprehensive
exposition of the approach, demonstrating that Pocock-
and O'Brien-Fleming-type boundaries can be recovered as
special cases of particular spending functions. The
Hwang-Shih-De Cani family of spending functions
[@hwang1990group] offered a further parameterization
that encompasses a wide range of boundary shapes,
including those more conservative than O'Brien-Fleming
and those more aggressive than Pocock.

The spending function framework is directly relevant to
the fully sequential paradigm because it can, in
principle, accommodate an arbitrarily large number of
analyses -- including the limiting case of continuous
monitoring. @betensky1998continuous explicitly
demonstrated the construction of a continuous stopping
boundary from an alpha spending function, providing the
mathematical machinery needed to implement fully
sequential monitoring within the modern spending
function framework. This result is central to our
reassessment: it implies that fully sequential
analysis with group size one is not a fundamentally
different methodology from group sequential analysis,
but rather the limiting case of the same framework.

## The Triangular Test

A parallel line of development was pursued by
@whitehead1983triangular, who proposed the triangular
test as a sequential procedure that allows stopping for
both efficacy and futility while maintaining
approximately optimal expected sample size properties.
@whitehead1997design extended this into a comprehensive
methodology for sequential clinical trials, including
the double triangular test for two-sided alternatives.
The triangular test occupies a conceptual middle ground
between fully sequential and group sequential methods:
it permits frequent interim analyses (though not
necessarily after every observation) and provides
natural boundaries for both early stopping directions.

@whitehead2013evolution provided a valuable historical
account of the evolution from Wald's SPRT through
Armitage's restricted procedures to the group sequential
and adaptive designs that dominate contemporary
practice. His account identifies the key practical
factors that drove the transition away from continuous
monitoring.

## Why the Field Moved to Group Sequential Methods

Several practical considerations explain why group
sequential methods eclipsed the fully sequential
approach despite the latter's theoretical efficiency
advantages:

1. **Multi-center trials.** As clinical trials grew
   larger and were conducted across multiple sites,
   it became logistically infeasible to update and
   evaluate the test statistic after every individual
   response. Data had to be collected, cleaned, and
   transmitted to a central coordinating center, a
   process that naturally imposed grouping.

2. **Delayed outcomes.** When the primary endpoint
   requires weeks or months to ascertain (e.g.,
   progression-free survival, functional recovery
   scores), continuous monitoring after each
   observation is not meaningful because many
   previously enrolled patients have not yet
   contributed their outcomes.

3. **Data monitoring committees.** The establishment
   of independent Data Safety Monitoring Boards
   (DSMBs) as a governance mechanism introduced a
   deliberative human element that requires
   scheduled meetings, not continuous algorithmic
   monitoring.

4. **Computational limitations.** In the 1970s and
   1980s, the computational burden of continuously
   updating boundaries and test statistics was
   non-trivial, though this constraint has been
   largely eliminated by modern computing.

5. **Regulatory practice.** Regulatory agencies
   became accustomed to reviewing trial designs with
   a small number of planned interim analyses,
   and guidance documents reflected this norm.

## Comparison of Fully Sequential and Group Sequential
   Approaches

@sebille2000comparison conducted a direct comparison of
four sequential methods -- the SPRT, the triangular test,
and one-parameter boundaries alongside Pocock and
O'Brien-Fleming group sequential boundaries -- and
evaluated their operating characteristics including
expected sample size, power, and type I error control.
@sebille2003sequential provided a more comprehensive
review contrasting strictly sequential methods with
group sequential designs, concluding that the
theoretical sample size savings of fully sequential
methods are partially offset by their practical
complexity.

@lai1980sequential established that certain sequential
stopping rules are asymptotically optimal from both
Bayesian and frequentist perspectives, providing
theoretical justification for continuous monitoring
when it is feasible. @xiong1996continuous extended the
sequential conditional probability ratio test (SCPRT)
framework to accommodate both continuous and group
sequential monitoring within a unified approach, further
demonstrating that the boundary between these paradigms
is more fluid than the dichotomy might suggest.

## Post-Estimation Challenges

A well-known complication of sequential analysis,
whether fully sequential or group sequential, is that
naive maximum likelihood estimates obtained at the
stopping time are biased. @emersonParameterEstimationFollowing1990
investigated parameter estimation following group
sequential tests and proposed orderings of the sample
space that yield improved point estimates and confidence
intervals. These estimation challenges are shared by
both paradigms, though the bias can be more pronounced
under continuous monitoring because the test statistic
is evaluated at a boundary-crossing point that is, by
construction, extreme.

## Modern Extensions: Longitudinal and Adaptive Designs

Recent work has extended group sequential methods to
richer data structures. @jeffriesLongitudinalClinicalTrials2015
considered longitudinal trials with adaptive choice of
follow-up time, @jeffriesDetectingTreatmentDifferences2018
incorporated covariate adjustment in group sequential
longitudinal studies, and
@jeffriesEvaluatingTreatmentEffects2023 further
generalized these methods to multivariate longitudinal
endpoints. These developments are relevant because they
demonstrate the continuing evolution of group sequential
methodology toward greater flexibility -- but they do
not address the question of whether reducing the group
size to one offers additional advantages.

The broader adaptive design literature, as codified
in @wassmer2016group and implemented in software such
as rpact [@wassmer2023rpact] and gsDesign
[@anderson2022gsdesign], has focused primarily on
sample size re-estimation, treatment selection, and
enrichment designs -- adaptations that presuppose a
group sequential monitoring structure.

## Safety Surveillance: A Revival of Continuous Monitoring

One domain in which fully sequential (or near-fully
sequential) monitoring has experienced a genuine
revival is post-market safety surveillance.
@kulkarni2016continuous compared continuous sequential
analysis using the maximized sequential probability
ratio test (maxSPRT) with group sequential methods for
vaccine safety monitoring. In this setting, the
practical constraints that drove the original move to
group sequential methods are less binding: data are
collected electronically in real time, the 'outcome'
(an adverse event) is typically ascertained quickly,
and the monitoring is algorithmic rather than
committee-based. This application demonstrates that
the fully sequential paradigm retains practical value
in settings where its original advantages -- minimal
expected sample size and rapid detection -- can be
realized.

## Present Study

We examine whether the fully sequential approach,
in which the stopping boundary is evaluated after
each individual response (group size equal to one),
merits reassessment for contemporary clinical trials
beyond safety surveillance. The question is
motivated by three developments since the 1970s:

1. Modern computing eliminates the historical
   computational barrier to continuous monitoring.
2. The alpha spending function framework
   [@lan1983discrete; @demetsInterimAnalysisAlpha1994]
   provides a rigorous theoretical foundation for
   continuous monitoring that controls type I error.
3. Electronic data capture and real-time data
   pipelines have reduced the logistical barriers
   to rapid data aggregation.

We investigate the operating characteristics of a fully
sequential design (group size one, alpha spending
function boundaries) relative to standard group
sequential designs with 3--5 interim analyses, focusing
on expected sample size, power, and the practical
trade-offs involved. The main-body comparison
(fixed versus K = 3, 5 versus fully sequential) largely
reproduces a result already established by
@sebille2000comparison, @sebille2003sequential, and
@jennison2000group (Chapter 7): most of the efficiency
gain of sequential monitoring is captured by a handful of
looks. The paper's own empirical contribution, developed
in the Appendix, is the K-grid analysis showing where
between K = 5 and K = N the returns actually diminish,
and the resulting case for high-frequency monitoring
(K = 20--50) as a practical middle ground; readers
mainly interested in that contribution can proceed
directly to the Appendix after the Methods and headline
Results below.

# Methods

## ADEMP structure

Following Morris, White, and Crowther [-@morris2019using], the
simulation is reported under the ADEMP framework.

- **Aims.** Compare the operating characteristics of fixed-sample,
  group-sequential (K = 3, 5), and fully sequential designs under
  Lan-DeMets O'Brien-Fleming spending across a range of true
  treatment effects.
- **Data-generating mechanism.** Two-arm continuous outcome with
  known variance; $n_{\max} = 200$ per arm in the main grid, and
  $n_{\max} = 85$ per arm in the power-calibrated sensitivity
  scenario (Results).
- **Estimands.** (i) The binary rejection decision at
  $\alpha = 0.05$; (ii) expected sample size at stopping; (iii) the
  naive MLE at stop, which exhibits early-stopping bias.
- **Methods.** Fixed design; group-sequential with K = 3, 5; fully
  sequential (K = n_max) in the main grid (Results). The finer
  K = 3, 5, 10, 20, 50, n_max grid is run only at $\delta = 0.5$, in
  both the appendix's overpowered scenario ($n_{\max} = 200$) and
  its power-calibrated companion ($n_{\max} = 85$).
- **Performance measures.** Rejection rate, mean and median N at
  stop, naive bias, and RMSE of the naive MLE, each reported with
  its Monte Carlo SE per Morris Table 6. For rejection-rate MCSE
  $\leq 1.1$ pp at $p = 0.5$, $n_{\mathrm{rep}} \geq 2{,}066$;
  $n_{\mathrm{rep}} = 2{,}000$ gives MCSE $\approx 1.12$ pp.

## Design Framework

We consider a two-arm parallel randomized trial with
a continuous primary endpoint and known variance. Under
the null hypothesis $H_0: \delta = 0$ and the
alternative $H_1: \delta = \delta_1$, we compare the
following monitoring strategies:

1. **Fixed-sample design:** No interim analysis;
   enroll $N$ subjects and analyze once.
2. **Group sequential design (K = 3):** Three equally
   spaced interim analyses with O'Brien-Fleming-type
   alpha spending.
3. **Group sequential design (K = 5):** Five equally
   spaced interim analyses with O'Brien-Fleming-type
   alpha spending.
4. **Fully sequential design (K = N):** Evaluate the
   boundary after each individual observation, using
   the Lan-DeMets alpha spending function
   corresponding to O'Brien-Fleming boundaries.

## Alpha Spending Function

For all group sequential and fully sequential designs,
we employ the Lan-DeMets spending function that
approximates the O'Brien-Fleming boundary:

$$\alpha^*(t) = 2 - 2\Phi\left(\frac{z_{\alpha/2}}{\sqrt{t}}\right)$$

where $t \in (0, 1]$ is the information fraction and
$\Phi$ is the standard normal CDF. This spending
function spends very little alpha early in the trial
and concentrates spending near the end, consistent
with the O'Brien-Fleming philosophy.

## Simulation Procedure

We evaluate the operating characteristics of each
design via Monte Carlo simulation:

- **True effect sizes:** $\delta \in \{0, 0.2, 0.5, 0.8\}$
  (standardized)
- **Maximum sample size:** $N = 200$ per arm for the
  main grid, plus a second scenario sized to give the
  fixed design approximately 90 percent power at
  $\delta = 0.5$ ($N = 85$ per arm; see Section
  "Power-Calibrated Sensitivity Scenario" below)
- **Significance level:** $\alpha = 0.05$ (two-sided)
- **Replications:** 2,000 per scenario (20,000 for the
  null scenario used in the type I error table; see
  below)
- **Random seed:** `set.seed(20260309)`, `RNGkind`
  `"L'Ecuyer-CMRG"`, set once

### Reproducibility

All simulation results reported in this manuscript are
generated by the standalone script
`analysis/scripts/run_all_simulations.R`, which sources
`analysis/scripts/sim_study.R`, sets the seed once, runs
every scenario the report uses, and writes versioned
`.rds` files to `analysis/data/derived_data/`. The Rmd
chunks below read those files with `readRDS()` and
perform no random sampling themselves, so the reported
numbers do not depend on knitr's chunk cache (which does
not replay RNG state across cached and uncached chunks
and previously left the appendix simulation's
reproducibility undocumented). To regenerate all results
from scratch:

```
Rscript analysis/scripts/run_all_simulations.R
```

```{r sim-source}
source(here::here("analysis", "scripts", "sim_study.R"))
derived_dir <- here::here(
  "analysis", "data", "derived_data"
)
sim_meta <- readRDS(
  file.path(derived_dir, "sim_metadata.rds")
)
```

```{r sim-run}
sim_results <- readRDS(
  file.path(derived_dir, "sim_results.rds")
)
sim_results_null <- readRDS(
  file.path(derived_dir, "sim_results_null.rds")
)
sim_results_90pow <- readRDS(
  file.path(derived_dir, "sim_results_90pow.rds")
)
```

# Results

## Headline findings

```{r headline-20, include=FALSE}
t1 <- sim_results_null
pow_tbl <- sim_results |> dplyr::filter(effect_size == 0.5)
```

Under the null ($\delta = 0$), evaluated at
$n_{\mathrm{rep}} = 20{,}000$ so the Monte Carlo SE is
precise enough to assess nominal coverage (MCSE
$\approx$
`r sprintf('%.4f', max(t1$mcse_rejection, na.rm = TRUE))`),
empirical type I error ranged from
`r sprintf('%.3f', min(t1$rejection_rate, na.rm = TRUE))`
to `r sprintf('%.3f', max(t1$rejection_rate, na.rm = TRUE))`
against nominal $\alpha = 0.05$, all designs within
roughly two Monte Carlo SEs of nominal. This supersedes
an earlier reading of this simulation at
$n_{\mathrm{rep}} = 2{,}000$, where the fully sequential
design's rejection rate (0.039) sat about 2.2 MCSEs
below nominal; at $n_{\mathrm{rep}} = 20{,}000$ the same
design's rejection rate is
`r sprintf('%.3f', t1$rejection_rate[t1$design == "Fully Seq"])`,
consistent with nominal control and indicating the
earlier shortfall was Monte Carlo noise rather than a
systematic conservatism of the K = N discrete spending
construction. At $\delta = 0.5$ and $N_{\max} = 200$
(the overpowered main-grid scenario; see the
power-calibrated scenario below for a realistic power
regime), the fully sequential design achieved rejection
rate
`r sprintf('%.3f', pow_tbl$rejection_rate[pow_tbl$design == 'Fully Seq'][1])`
with mean sample size
`r sprintf('%.1f', pow_tbl$mean_n[pow_tbl$design == 'Fully Seq'][1])`
(MCSE
`r sprintf('%.2f', pow_tbl$mcse_mean_n[pow_tbl$design == 'Fully Seq'][1])`),
versus the fixed design's sample size of
`r sprintf('%.0f', pow_tbl$mean_n[pow_tbl$design == 'Fixed'][1])`
at rejection rate
`r sprintf('%.3f', pow_tbl$rejection_rate[pow_tbl$design == 'Fixed'][1])`.
Monte Carlo SEs (MCSE) are reported alongside every
performance measure in the tables below.

## Type I Error Control

```{r type1-table}
library(knitr)
library(kableExtra)

type1 <- sim_results_null |>
  dplyr::select(design, rejection_rate, mcse_rejection,
    mean_n, mcse_mean_n, median_n)

kable(
  type1,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  col.names = c("Design", "Type I Error", "MCSE",
    "Mean N", "MCSE(N)", "Median N"),
  caption = "Type I error rates and expected sample
    sizes under the null hypothesis ($\\delta = 0$,
    $n_{\\mathrm{rep}} = 20{,}000$).",
  digits = 4
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 10
  )
```

## Power and Expected Sample Size

```{r power-table}
power_tab <- sim_results |>
  dplyr::filter(effect_size > 0) |>
  dplyr::select(design, effect_size, rejection_rate,
    mcse_rejection, mean_n, mcse_mean_n, median_n)

kable(
  power_tab,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  col.names = c("Design", "$\\delta$", "Power", "MCSE",
    "Mean N", "MCSE(N)", "Median N"),
  caption = "Power and expected sample sizes under
    alternative hypotheses ($N_{\\max} = 200$ per arm;
    power is at or near ceiling for $\\delta \\geq 0.5$,
    see the power-calibrated scenario below).",
  digits = 3
) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 10
  )
```

## Power-Calibrated Sensitivity Scenario

The main grid above is conducted at $N_{\max} = 200$,
which gives the fixed design essentially saturated power
(0.998--1.000) at $\delta = 0.5$ and $0.8$. Sample size
savings computed in that regime are conditional on gross
overpowering and do not establish how much benefit
sequential monitoring offers when a trial is sized
conventionally. We add a second scenario, sized so the
fixed design has approximately 90 percent power at
$\delta = 0.5$: $N_{\max} =$
`r sim_meta$n90` per arm, from the two-sample normal mean
formula $N = 2(z_{\alpha/2} + z_\beta)^2 / \delta^2$.

```{r power90-table}
pow90_tab <- sim_results_90pow |>
  dplyr::select(design, rejection_rate, mcse_rejection,
    mean_n, mcse_mean_n, median_n)

kable(
  pow90_tab,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  col.names = c("Design", "Power", "MCSE", "Mean N",
    "MCSE(N)", "Median N"),
  caption = paste0(
    "Power and expected sample size at $\\delta = 0.5$",
    ", $N_{\\max} = ", sim_meta$n90,
    "$ per arm (fixed design at approximately 90\\%",
    " power)."
  ),
  digits = 3
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 10
  )
```

In this realistically powered regime, the fully
sequential design's power drops to
`r sprintf('%.3f', pow90_tab$rejection_rate[pow90_tab$design == "Fully Seq"])`,
about
`r sprintf('%.1f', 100 * (pow90_tab$rejection_rate[pow90_tab$design == "Fixed"] - pow90_tab$rejection_rate[pow90_tab$design == "Fully Seq"]))`
percentage points below the fixed design's
`r sprintf('%.3f', pow90_tab$rejection_rate[pow90_tab$design == "Fixed"])`,
while still saving
`r sprintf('%.1f', 100 * (1 - pow90_tab$mean_n[pow90_tab$design == "Fully Seq"] / pow90_tab$mean_n[pow90_tab$design == "Fixed"]))`
percent of expected sample size relative to the fixed
design. The sample size savings are real but smaller,
and the power cost of sequential stopping is visible,
in this regime than in the overpowered main-grid
scenario above.

The same power cost is visible, if less dramatically,
even in the overpowered main grid at
$\delta = 0.2$, the smallest nonzero effect size
evaluated: fixed-design power is
`r sprintf('%.3f', sim_results$rejection_rate[sim_results$design == "Fixed" & sim_results$effect_size == 0.2])`
versus
`r sprintf('%.3f', sim_results$rejection_rate[sim_results$design == "Fully Seq" & sim_results$effect_size == 0.2])`
for the fully sequential design, a loss that cuts
against the paper's ASN-savings thesis and is worth
weighing against the sample size reduction the same
design offers at larger effect sizes.

## Expected Sample Size by Effect Size

```{r asn-plot, fig.cap="Expected sample size (ASN) as a function of the true standardized effect size across four monitoring designs. The fully sequential design achieves the lowest ASN for large effects but offers diminishing returns relative to the five-look group sequential design."}
library(ggplot2)

sim_results |>
  ggplot(aes(x = effect_size, y = mean_n,
    linetype = design, shape = design)) +
  geom_line() +
  geom_point(size = 2.5) +
  scale_linetype_manual(
    values = c("solid", "dashed", "dotted",
      "twodash"),
    labels = c("Fixed", "GS (K=3)", "GS (K=5)",
      "Fully Sequential")
  ) +
  scale_shape_manual(
    values = c(16, 17, 15, 4),
    labels = c("Fixed", "GS (K=3)", "GS (K=5)",
      "Fully Sequential")
  ) +
  labs(
    x = "Standardized Effect Size",
    y = "Expected Sample Size (per arm)",
    linetype = "Design",
    shape = "Design"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

## Stopping Time Distribution

Stopping fractions are point masses at each design's K
analysis times (plus a mass at 1.0 for trials that reach
the final look, including every fixed-design trial by
construction). A kernel density smooths these point
masses into a spread that does not exist in the data; we
plot the empirical CDF instead, which represents a
discrete stopping-time distribution honestly.

```{r stopping-ecdf, fig.cap="Empirical cumulative distribution of stopping times (information fraction at termination) under a moderate effect size (delta = 0.5). Each design's stopping fraction is a step function with mass only at its own analysis times; the fully sequential design's steps are finer-grained than the group sequential designs'."}
stop_df <- sim_results |>
  dplyr::filter(effect_size == 0.5) |>
  dplyr::select(design, stop_fractions) |>
  tidyr::unnest_longer(stop_fractions,
    values_to = "stop_fraction")

ggplot(stop_df, aes(x = stop_fraction,
  linetype = design)) +
  stat_ecdf(geom = "step", linewidth = 0.6) +
  scale_linetype_manual(
    values = c("solid", "dashed", "dotted",
      "twodash"),
    labels = c("Fixed", "GS (K=3)", "GS (K=5)",
      "Fully Sequential")
  ) +
  labs(
    x = "Information Fraction at Stopping",
    y = "Cumulative Probability",
    linetype = "Design"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

# Discussion

We shall interpret the results in the context of both
the theoretical and practical dimensions of clinical
trial monitoring.

From a purely statistical perspective, the fully
sequential design (group size one) represents the
limiting case of the group sequential framework and,
as predicted by the classical theory
[@wald1947sequential; @armitage1975sequential], achieves
the lowest expected sample size when the true treatment
effect is large. However, the marginal gain from
moving from five interim analyses to continuous
monitoring is modest, consistent with the well-known
result that most of the efficiency benefit of sequential
monitoring is captured by the first few interim
analyses [@jennison2000group, Chapter 7].

The alpha spending function framework
[@lan1983discrete; @demetsInterimAnalysisAlpha1994]
provides the theoretical foundation that makes fully
sequential monitoring rigorous within the modern
inferential framework. @betensky1998continuous
demonstrated that continuous stopping boundaries can
be constructed directly from spending functions. At
$n_{\mathrm{rep}} = 20{,}000$ (precise enough to
distinguish nominal control from a few-thousandths
deviation), all four designs' empirical type I error
rates sit within roughly two Monte Carlo SEs of the
nominal $\alpha = 0.05$, supporting the claim that type
I error is controlled at the nominal level under
continuous monitoring as constructed here (per-pair
looks with rounded analysis times); we do not evaluate
whether every discrete-K approximation to a continuous
spending function shares this property.

The designs compared here stop only on rejection of the
null; no futility (beta-spending) boundary is
implemented, so a trial that fails to cross the upper
boundary continues to $N_{\max}$. Consequently, under
$H_0$ the mean sample size is 197.7 to 200 for every
design (Table with the type I error results above): the
simulation demonstrates no expected sample size benefit
under the null, even though the Introduction motivates
fully sequential monitoring partly through Armitage's
lower futility and harm boundary, and the appendix's
architecture proposal recommends conditional-power
futility alerts. The ASN savings reported throughout
this report are therefore savings from efficacy-only
stopping under a true, nonzero treatment effect; they
say nothing about the sample size benefit -- if any --
of stopping early for futility, which would require
adding a beta-spending boundary and is left for future
work (see Future Research). Readers should read every
claim of sample size "savings" in this report as
conditional on that scope.

Every simulated look is evaluated after a treated-control
pair (K = 200 corresponds to 200 pairs, 400 subjects, at
the main-grid $N_{\max}$), not after each individual
response as a literal reading of "fully sequential" (group
size one, following one observation at a time) might
suggest. We retain the paired framing throughout because
the two-arm normal DGM pairs enrollment across arms by
construction; the qualitative comparison to group
sequential designs with K = 3 or K = 5 looks is
unaffected, but the label "fully sequential" here means
"one look per enrolled pair," not "one look per
observation."

The practical arguments against fully sequential
monitoring -- multi-center logistics, delayed outcomes,
DSMB governance -- remain valid for most large
confirmatory trials. However, several evolving features of contemporary
clinical trial practice may narrow the practical gap:

- Electronic data capture systems increasingly
  support near-real-time data availability.
- Decentralized trial designs reduce the data
  transmission delays inherent in multi-site studies.
- Rapidly ascertained endpoints (e.g., biomarker
  responses, short-term safety events) eliminate the
  delayed-outcome objection for certain trial types.
- Algorithmic monitoring can complement, rather than
  replace, DSMB review -- a continuous monitoring
  system could trigger an ad hoc DSMB meeting when
  evidence crosses a threshold.

The revival of continuous sequential monitoring in
post-market safety surveillance
[@kulkarni2016continuous] demonstrates that the
approach is already viable in settings that share
some of these characteristics. Whether the same logic
extends to confirmatory efficacy trials is an open
question that depends on the specific endpoint,
timeline, and regulatory context.

The comparison of sequential approaches by
@sebille2000comparison and @sebille2003sequential
supports the conclusion that the choice between
fully sequential and group sequential monitoring is
driven more by practical logistics than by
statistical efficiency. When the practical barriers
to continuous monitoring are low, the fully sequential
approach offers real, if modest, sample size savings.
When those barriers are high, the group sequential
approach sacrifices little efficiency while gaining
substantial operational simplicity.

## Position Relative to Anytime-Valid Inference

The spending-function approach evaluated here is not the
only contemporary route to continuous monitoring with
error control. Since roughly 2011, a distinct literature
on anytime-valid (also called always-valid) inference has
addressed the same problem -- valid inference at
arbitrary, possibly data-dependent stopping times -- from
a different starting point, and is deployed at
production scale in industrial A/B testing
[@johari2017peeking]. Confidence sequences
[@robbins1970statistical; @howard2021time] provide
intervals that are simultaneously valid at every sample
size, so a user may stop for any reason, including
looking at every new observation, without inflating type
I error; e-values and e-processes generalize this to a
broader game-theoretic statistics framework
[@ramdas2023game]. Kulldorff and colleagues'
maximized sequential probability ratio test
[@kulldorff2011maximized] is the seminal continuous-
monitoring method for drug and vaccine safety
surveillance and predates the safety-surveillance
application cited earlier [@kulkarni2016continuous].
The classical monitoring literature this report draws on
is thoroughly synthesized in @proschan2006statistical,
and the repeated confidence interval approach of
@jennison1989interim, referenced without citation in
the appendix's software architecture proposal, is the
group sequential analogue of a confidence sequence.

These methods differ from the Lan-DeMets alpha-spending
construction used throughout this report primarily in
how they are derived (test-martingale or confidence-
sequence arguments versus information-fraction spending
functions) and in their typical use case (continuous,
often unbounded monitoring in digital experimentation
versus a fixed-$N_{\max}$, discretely monitored clinical
trial). We do not implement an mSPRT or confidence-
sequence comparator in the simulation reported here; a
head-to-head comparison of expected sample size and
naive-estimator bias between a K = 20-50 spending-
function design and a comparably tuned confidence
sequence is a natural extension and is listed under
Future Research. Absent that comparison, we do not claim
that the spending-function approach dominates
anytime-valid alternatives, only that it remains the more
familiar framework for regulated confirmatory trials and
that the high-frequency monitoring (K = 20 to 50)
direction proposed in the appendix is one practical way
to narrow the gap between the K = 3-5 convention and the
anytime-valid literature that most trialists do not yet
use.

## Limitations

This analysis considers only a two-arm trial with a
continuous endpoint and known variance. The comparison
may differ for binary endpoints, time-to-event
outcomes, or designs with unknown variance. The
simulation uses a fixed maximum sample size; in
practice, the maximum may itself be adapted. The
post-estimation bias issue
[@emersonParameterEstimationFollowing1990] is
acknowledged but not directly evaluated here. No
futility (beta-spending) boundary is implemented; all
reported ASN savings are efficacy-only-stopping savings
(see Discussion). Each design-by-effect-size cell in the
main simulation grid draws independent normal outcome
data rather than reusing a common random-number stream
across designs (`run_simulation()` in
`analysis/scripts/sim_study.R` calls
`simulate_one_trial()` independently for every design);
cross-design contrasts at a fixed effect size are
therefore somewhat noisier than a common-random-numbers
design would produce, though each design's own MCSE is
correctly computed from its own replications. Finally,
the main-grid comparison at $N_{\max} = 200$ is
overpowered for $\delta \geq 0.5$; we address this with
the power-calibrated sensitivity scenario in Results, but
do not rerun the appendix's full K-grid diminishing-
returns curve at every effect size in the properly
powered regime, only at $\delta = 0.5$ (see Appendix).

## Data and Code Availability

All simulation code, the manuscript source, and this
report's compiled output are versioned in the git
repository containing this document
(`sequentialanalysis`). Simulation results are generated
by `analysis/scripts/run_all_simulations.R`, which is
deterministic given the seed recorded in
`analysis/data/derived_data/sim_metadata.rds`
(`set.seed(20260309)`, `RNGkind("L'Ecuyer-CMRG")`) and
writes the `.rds` files this report reads. Key package
versions used to generate the results in this revision,
recorded at simulation time: R
`r sim_meta$r_version`; gsDesign
`r sim_meta$package_versions[["gsDesign"]]`; dplyr
`r sim_meta$package_versions[["dplyr"]]`; ggplot2
`r sim_meta$package_versions[["ggplot2"]]`; knitr
`r sim_meta$package_versions[["knitr"]]`. The full
dependency lockfile is `renv.lock` at the repository
root.

# Future Research

1. **Extension to time-to-event endpoints.** The
   fully sequential paradigm is naturally suited to
   event-driven analyses; the connection between
   continuous monitoring and the calendar-time
   spending function [@lan1983discrete] warrants
   formal evaluation in the survival setting.

2. **Integration with adaptive sample size
   re-estimation.** Combining continuous monitoring
   with unblinded sample size re-estimation
   [@wassmer2016group] could yield designs that
   adapt both the monitoring frequency and the
   maximum sample size as data accumulate.

3. **Bayesian continuous monitoring.** The Bayesian
   sequential framework offers a natural continuous
   monitoring approach via posterior probabilities.
   A formal comparison of Bayesian continuous
   monitoring with frequentist fully sequential
   designs would clarify the relative advantages.

4. **Software development.** While gsDesign
   [@anderson2022gsdesign] and rpact
   [@wassmer2023rpact] support many interim analyses,
   dedicated tools for fully sequential (K = N)
   design with spending function boundaries and
   bias-adjusted estimation would facilitate
   practical adoption.

5. **Regulatory considerations.** Engagement with
   regulatory agencies on the acceptability of
   fully sequential designs, particularly for trials
   with rapidly ascertained endpoints, is a necessary
   step before broader adoption.

6. **Post-estimation methods.** Development and
   evaluation of bias-corrected estimators and
   confidence intervals specifically tailored to
   the fully sequential setting, building on
   Emerson and Fleming
   [-@emersonParameterEstimationFollowing1990].

7. **Decentralized trials.** The growing use of
   decentralized trial platforms with real-time data
   capture creates a natural testbed for evaluating
   fully sequential monitoring in practice.

8. **Comparison against anytime-valid inference.** A
   head-to-head simulation of the spending-function
   designs evaluated here against a mixture sequential
   probability ratio test or a confidence-sequence
   procedure [@robbins1970statistical; @howard2021time;
   @ramdas2023game], matched on type I error and power,
   would establish whether the high-frequency
   spending-function designs proposed in the appendix
   remain competitive with the anytime-valid literature
   on ASN and estimator bias.

9. **Futility stopping.** Add a beta-spending futility
   boundary (e.g., `gsDesign` `test.type = 4`) to each
   design and re-evaluate ASN under the null and under
   small effect sizes, which the efficacy-only designs
   evaluated in this report do not capture.

\newpage
\appendix

# Appendix: High-Frequency Monitoring as a Practical
  Middle Ground

## Motivation

The main body of this report frames the comparison as
fully sequential (K = N) versus the standard group
sequential paradigm (K = 3--5). The simulation results
and the historical literature converge on the same
conclusion: the statistical efficiency gains from
continuous monitoring are real but modest, while the
practical barriers remain substantial. This framing,
however, presents a false dichotomy. The alpha spending
function framework imposes no constraint on K. The
question is not whether to use K = 5 or K = N, but
rather: what value of K captures most of the efficiency
benefit while remaining operationally tractable?

We propose that the most productive direction for
sequential analysis in 2026 is not a revival of the
fully sequential paradigm per se, but rather the
development of high-frequency monitoring designs with
K in the range of 20--50, supported by three pillars:
(i) software that handles large K without numerical
instability, (ii) bias-corrected estimation tailored
to the high-frequency setting, and (iii) automated
alerting systems that translate continuous statistical
monitoring into actionable triggers for human review.

## Diminishing Returns in K: An Empirical Analysis

To motivate the choice of K, we extend the simulation
from the main report to a finer grid of monitoring
frequencies. We fix the effect size at $\delta = 0.5$
and vary K across 3, 5, 10, 20, 50, and N = 200,
all under the Lan-DeMets O'Brien-Fleming spending
function.

```{r hf-sim}
# hf_results is generated by analysis/scripts/run_all_simulations.R
# (single seed, no chunk cache; see Simulation Procedure/
# Reproducibility above) and read here from the versioned .rds file.
hf_results <- readRDS(
  file.path(derived_dir, "hf_results.rds")
)
hf_results_90pow <- readRDS(
  file.path(derived_dir, "hf_results_90pow.rds")
)
```

```{r hf-table}
hf_display <- hf_results |>
  dplyr::select(
    label, rejection_rate, mcse_rejection, mean_n,
    mcse_mean_n, median_n, sd_n, bias_naive,
    mcse_bias_naive, rmse_naive, mcse_rmse_naive,
    pct_saving
  )

kable(
  hf_display,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  digits = 3,
  col.names = c(
    "Design", "Power", "MCSE", "Mean N", "MCSE(N)",
    "Median N", "SD(N)", "Bias", "MCSE(Bias)", "RMSE",
    "MCSE(RMSE)", "\\% Saved"
  ),
  caption = paste(
    "Operating characteristics across monitoring",
    "frequencies ($\\delta = 0.5$, $N_{\\max} = 200$",
    "per arm, $n_{\\mathrm{rep}} = 2{,}000$). Bias and",
    "RMSE refer to the naive MLE at the stopping time.",
    "Percent saved is relative to the fixed-sample",
    "maximum. This scenario is overpowered (power",
    "0.996-1.000 throughout); see the properly powered",
    "companion table below."
  )
) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 8
  )
```

```{r hf-table-90pow}
hf90_display <- hf_results_90pow |>
  dplyr::select(
    label, rejection_rate, mcse_rejection, mean_n,
    mcse_mean_n, bias_naive, mcse_bias_naive,
    pct_saving
  )

kable(
  hf90_display,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  digits = 3,
  col.names = c(
    "Design", "Power", "MCSE", "Mean N", "MCSE(N)",
    "Bias", "MCSE(Bias)", "\\% Saved"
  ),
  caption = paste0(
    "Operating characteristics across monitoring",
    " frequencies at $\\delta = 0.5$, $N_{\\max} = ",
    sim_meta$n90,
    "$ per arm (fixed design at approximately 90\\%",
    " power). Compare with the overpowered table above:",
    " power now visibly declines with K (0.897 at K=3 to",
    " 0.891 at K=", sim_meta$n90,
    "), and the percent-saved curve is smaller in",
    " absolute terms at every K."
  )
) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 9
  )
```

At the properly powered $N_{\max} = `r sim_meta$n90`$
scenario, the diminishing-returns shape survives: most of
the achievable savings are again captured by K = 20, but
the absolute savings are smaller (K = 20 saves
`r sprintf('%.1f', hf_results_90pow$pct_saving[hf_results_90pow$k == 20])`
percent versus fixed, compared with
`r sprintf('%.1f', hf_results$pct_saving[hf_results$k == 20])`
percent in the overpowered $N_{\max} = 200$ scenario),
and the power cost of frequent monitoring, invisible when
power is at ceiling, is now measurable: power falls from
`r sprintf('%.3f', hf_results_90pow$rejection_rate[hf_results_90pow$k == 3])`
at K = 3 to
`r sprintf('%.3f', hf_results_90pow$rejection_rate[hf_results_90pow$label == "Fully Seq"])`
at fully sequential monitoring.

```{r hf-asn-plot, fig.cap="Expected sample size and sample size savings as a function of K (number of planned analyses) under delta = 0.5, n-max = 200 (overpowered scenario). The curve flattens rapidly: relative to the fixed design, K = 20 captures about 96 percent of the ASN reduction achieved by fully sequential monitoring (52.8 over 55.0); relative to K = 3, K = 20 captures about 87 percent of the incremental reduction from K = 3 to fully sequential, that is, (52.8 minus 38.7) over (55.0 minus 38.7). See the properly powered companion scenario above for the same comparison outside the ceiling-power regime.", fig.height=3.5}
library(ggplot2)

p1 <- ggplot(hf_results,
  aes(x = k, y = mean_n)) +
  geom_line() +
  geom_point(size = 2.5, shape = 16) +
  scale_x_continuous(
    breaks = c(3, 5, 10, 20, 50, 200),
    trans = "log10",
    labels = c("3", "5", "10", "20", "50", "200")
  ) +
  labs(
    x = "Number of Analyses (K)",
    y = "Expected Sample Size (per arm)"
  ) +
  theme_minimal(base_size = 10)

p2 <- ggplot(hf_results,
  aes(x = k, y = pct_saving)) +
  geom_line() +
  geom_point(size = 2.5, shape = 17) +
  scale_x_continuous(
    breaks = c(3, 5, 10, 20, 50, 200),
    trans = "log10",
    labels = c("3", "5", "10", "20", "50", "200")
  ) +
  labs(
    x = "Number of Analyses (K)",
    y = "Sample Size Savings (%)"
  ) +
  theme_minimal(base_size = 10)

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

The key result is the shape of the curve. Moving from
K = 3 to K = 10 captures the majority of the available
efficiency gain. Moving from K = 10 to K = 20 adds a
meaningful but smaller increment. Beyond K = 20, the
returns diminish sharply. By K = 50, the design is
operationally indistinguishable from fully sequential
monitoring for practical purposes, yet the
infrastructure demands are far more manageable.

## Bias Under High-Frequency Monitoring

```{r hf-bias-plot, fig.cap="Naive MLE bias at the stopping time as a function of K. More frequent monitoring increases the conditional bias at stopping because the test statistic is more likely to cross the boundary at an extreme point. The bias remains small in absolute terms and tends to increase with K, within Monte Carlo noise (see MCSE(Bias) in the table above).", fig.height=3}
ggplot(hf_results,
  aes(x = k, y = bias_naive)) +
  geom_line() +
  geom_point(size = 2.5, shape = 15) +
  geom_hline(yintercept = 0, linetype = "dashed",
    color = "grey50") +
  scale_x_continuous(
    breaks = c(3, 5, 10, 20, 50, 200),
    trans = "log10",
    labels = c("3", "5", "10", "20", "50", "200")
  ) +
  labs(
    x = "Number of Analyses (K)",
    y = "Bias of Naive MLE"
  ) +
  theme_minimal(base_size = 10)
```

A well-known consequence of optional stopping is that
the naive maximum likelihood estimate at the stopping
time is biased away from the null
[@emersonParameterEstimationFollowing1990]. The figure
above shows that this bias tends to increase with K
(0.046 at K = 50 versus 0.044 at K = 200, a reversal
within Monte Carlo noise: MCSE(Bias) is on the order of
0.003-0.004 at these K, so strict monotonicity is not
established): more frequent monitoring creates more
opportunities for the test statistic to cross the
boundary at an extreme realization. However, two
observations temper this concern. First, the absolute
magnitude of the bias remains modest even at K = 200.
Second, the bias is
a known and correctable phenomenon. Several approaches
to bias-corrected estimation following sequential
monitoring have been developed, including median
unbiased estimation, conditional MLE methods, and
bootstrap-based corrections. Practical adoption of
high-frequency monitoring will require that these
methods be packaged into accessible software rather
than remaining the province of methodological
specialists.

## A Software Architecture for High-Frequency
   Monitoring

The current generation of group sequential software
-- gsDesign [@anderson2022gsdesign], rpact
[@wassmer2023rpact], and the Lan-DeMets boundary
programs -- was designed for the K = 3--5 regime. While
these tools can technically accommodate larger K, they
were not optimized for it. A purpose-built system for
high-frequency monitoring would require three
components:

**Boundary computation engine.** The spending function
must be evaluated at K = 20--50 information fractions,
and the corresponding critical boundaries computed via
numerical integration or recursive application of the
multivariate normal distribution. For moderate K
(up to 50), existing algorithms based on the work of
@jennison2000group scale adequately. For K approaching
N, direct computation becomes prohibitive and the
continuous boundary approximation of
@betensky1998continuous should be used instead. A
practical implementation would pre-compute boundaries
for a fine grid of information fractions and
interpolate at the actual analysis times.

**Bias-corrected estimation module.** At each interim
analysis, the system should report not only the naive
test statistic and p-value but also:

- The median unbiased estimate of the treatment effect,
  computed by inverting the stage-wise ordering of the
  sample space.
- A repeated confidence interval that maintains
  simultaneous coverage across all analysis times.
- The conditional power given the data observed so
  far, to inform DSMB deliberations about futility.

These quantities are computationally more demanding
than the naive estimates but are feasible for K up to
50 with modern hardware.

**Automated alerting layer.** The most consequential
practical innovation would be an alerting system that
decouples statistical monitoring from DSMB scheduling.
In the current paradigm, the DSMB meets at
pre-specified calendar times, and the interim analysis
is conducted at that meeting. In a high-frequency
monitoring paradigm, the statistical comparison would
run continuously (or at each new data lock), and the
system would generate alerts at three thresholds:

1. *Efficacy signal* -- the test statistic crosses the
   upper spending function boundary. This triggers an
   expedited DSMB meeting to review the unblinded data
   and consider early termination.
2. *Futility signal* -- the conditional power falls
   below a pre-specified threshold (e.g., 10--20\%).
   This triggers a DSMB review of whether to continue
   enrollment.
3. *Safety signal* -- a secondary monitoring rule for
   adverse events crosses a safety boundary. This
   triggers immediate safety review regardless of the
   efficacy analysis schedule.

The DSMB retains full authority over all stopping
decisions. The system merely ensures that evidence is
not accumulating unexamined between scheduled meetings.
This is particularly relevant for trials in which
enrollment is fast relative to the DSMB meeting
schedule: if 50 patients enroll between quarterly
meetings, a group sequential design with K = 5 may
miss the opportunity to stop the trial at the
statistically optimal time.

## The Diminishing Marginal Cost of More Analyses

A counterargument to high-frequency monitoring is that
each interim analysis incurs administrative cost: data
locks, database reconciliation, statistical programming,
and DSMB coordination. This is true in the current
paradigm but largely an artifact of it. If the
monitoring system is automated -- data are ingested
continuously from electronic data capture, boundaries
are pre-computed, and the test statistic is updated
programmatically -- then the marginal cost of the 21st
analysis is approximately zero. The cost structure
shifts from per-analysis overhead to a fixed upfront
investment in infrastructure. For large trials where
the cost per enrolled patient is high and the potential
savings from early termination are substantial, this
investment is easily justified.

## Practical Recommendations

Based on the simulation results and the considerations
above, we offer the following concrete recommendations:

1. **Default to K = 20 rather than K = 5.** For trials
   with electronically captured endpoints and
   turnaround times of less than one month, the
   spending function framework supports 20 planned
   analyses with no loss of type I error control and
   meaningful gains in expected sample size. The
   practical overhead is minimal if the monitoring
   system is automated.

2. **Use the O'Brien-Fleming spending function.** At
   high K, the O'Brien-Fleming spending function
   spends negligible alpha at early analyses, so the
   cost of 'looking more often' is almost entirely
   borne by a trivially small increase in the maximum
   sample size (typically less than 1\% inflation
   relative to the fixed-sample design).

3. **Report bias-corrected estimates at every
   analysis.** The naive MLE should never be the sole
   estimate presented to a DSMB. At minimum, report
   the median unbiased estimate and the repeated
   confidence interval alongside the naive values.

4. **Build the alerting system first.** The most
   impactful near-term investment is not in novel
   statistical methodology but in software
   engineering: a system that ingests data, updates
   the test statistic, evaluates it against
   pre-computed boundaries, and generates structured
   alerts when thresholds are crossed. This system
   should produce a standardized DSMB report at each
   triggered analysis, including the spending function
   status (cumulative alpha spent), the bias-corrected
   estimate, conditional power, and a recommendation
   for whether a DSMB meeting is warranted.

5. **Treat K as a design parameter to be optimized.**
   Rather than choosing K by convention (the current
   default of 3--5), select K by explicitly trading
   off expected sample size savings against
   infrastructure cost. The diminishing-returns curve
   from the simulation provides the empirical basis
   for this trade-off.

# References

A remediation plan is recorded in the audit file and will be
taken up in a subsequent revision.
